---
title: "Homework 2"
author: "Solution"
date: "Due 6 February 2020, 11:59pm"
output:
  pdf_document:
    includes:
      in_header: ../support/432macros.tex
    number_sections: no
    template: ../support/dm-docs.tex
    toc: no
  html_document:
    theme: flatly
    fig_caption: yes
---

```{r, include=FALSE}
# General set-up for the report:
# Don't print out code
# Save results so that code blocks aren't re-run unless code
# changes (cache), _or_ a relevant earlier code block changed (autodep),
# don't clutter R output with messages or warnings (message, warning)
library(MASS)
library(knitr)
library(tidyverse)
opts_chunk$set(echo=FALSE, fig.align = TRUE, fig.height = 4, fig.width = 6,
               cache=TRUE, autodep=TRUE,
               message=FALSE, warning=FALSE)
# Turn off meaningless clutter in summary() output
options(show.signif.stars=FALSE)
```

## Introduction

Appraising residential real estate --- predicting the price at which it could
be sold, under current market conditions --- is important not only for people
buying and selling houses to live in, but also for real estate developers,
mortgage lenders, and local tax assessors.  Currently, appraisal is usually
done by skilled professionals who make a good living at it, so naturally there
is interest in replacing them by machines.  In this report, we investigate the
feasibility of real estate appraisal by means of linear statistical models.

Specific points of interest to the client include the relationship between the
quality of the house's construction and its price; the relationship between age
and price, and whether this changes depending on proximity to a highway; and
the relationship between price, the finished area of the house, and the number
of bedrooms.

## Exploratory data analysis

```{r, include=FALSE}
# Load the data
real_estate <- read_csv(
  "http://www.stat.cmu.edu/~cshalizi/mreg/15/hw/08/real-estate.csv",
  col_types = "dddiififdfdf") %>% 
  mutate(
    Quality = fct_recode(Quality, High = "1", Medium = "2", Low = "3"),
    AC = fct_recode(Airconditioning, Yes = "1", No = "0"),
    Pool = fct_recode(Pool, Yes="1", No="0"),
    Highway = fct_recode(as.factor(AdjHighway), Adjacent="1", Not = "0")
  )

```

The data, supplied by an undisclosed client, come from a selection of
"arms-length" residential real estate transactions in an unnamed city in the
American midwest in 2002.  This records, for `r nrow(real_estate)`
transactions, the sale price of the house, its finished area and the area of
the lot, the number of bedrooms, the number of bathrooms, the number of cars
that will fit in its garage, the year it was built, whether it has air
conditioning, whether it has a pool, whether it is adjacent to a highway, and
the quality of construction, graded from low to medium or high.  It is notable
that, except for highway adjacency, we have no information about the location
of the houses, though this is proverbially a very important influence on their
price, through access to schools, commuting time, land value, etc.

```{r, pairsplot, fig.width=8, fig.height=8, fig.cap="Pairs plot for quantitative variables"}
# Base R version
# pairs(Price ~ Sqft+Bedroom+Bathroom+Garage+YearBuild+Lot, data=real.estate,
#      pch=19, cex=0.3)

# Pretty version
library(GGally)
library(cowplot)
ggpairs(
  real_estate, 
  c("Price", "Sqft", "Bedroom","Bathroom","Garage", "YearBuild", "Lot"),
  upper="blank", 
  lower = list(continuous = wrap("points", alpha=.4, color="purple"))
  ) + theme_cowplot(8)
```


Pairwise scatter-plots for the quantitative variables (Figure 1) show that,
unsurprisingly, there is a positive relationship between price and area
(stronger for finished area than the total lot size), and price and the number
of bedrooms, bathrooms, or garage slots (all three of which are strongly
positively related to each other).  The relation between price and these three
"count" variables could well be linear.  There is a positive relation between
price and the year of construction, i.e., newer houses cost more.  Newer houses
also tend to be larger, both in finished area and the number of rooms, though
not to have bigger lots.

Inspection of the plots shows there is one record with 0 bedrooms, 0 bathrooms,
and a three-car garage with air conditioning.  This is either not a piece of
residential real estate, or its data is hopelessly corrupt; either way, we drop
it from the data from now on.

```{r}
# What's that outlier with zero bedrooms and zero bathrooms?
  # From visual inspection of pairs plot, no other weirdness with houses without
  # bedrooms or bathrooms
# bad.row <- which(real.estate$Bedroom==0 & real.estate$Bathroom==0)
# Examination of real.estate[bad.row,] shows Garage==3, Airconditioning==1, etc.
# Remove the bad row
real_estate <- dplyr::filter(real_estate, !(Bedroom == 0 & Bathroom==0))
```


```{r, price-given-discrete-predictors-gg, fig.cap="Conditional distributions of price given qualitative predictors. Box widths reflect the number of points in each group, notches show medians plus/minus a margin of error."}
real_estate %>% 
  dplyr::select(Price, AC, Pool, Highway, Quality) %>% 
  pivot_longer(-Price) %>%
  ggplot(aes(y=Price/10000, x=value, fill=value)) + geom_boxplot() + 
  facet_wrap(~name, ncol=2, scales='free_x') +
  theme_cowplot(12) + theme(legend.position = 'none') + ylab("Price ($10000s)") + 
  scale_fill_brewer(palette = "Set1")
```


Box-plots, showing the conditional distribution of price for each level of the
categorical predictors, suggest that houses with air-conditioning and pools are
more expensive, that being next to a highway makes little difference, and that
higher quality of construction implies, on average, higher prices.  The
mid-points of the boxes for quality don't _quite_ fall on a straight line, so
treating quality as a numerical variable isn't obviously compelling, but not
clearly crazy either.

## Initial Modeling

To answer the client's questions, our model should include quality, finished
area, the number of bedrooms (and the interaction between those two), and the
year the house was built and whether it is adjacent to a highway (and the
interaction between those two).  Based on our EDA, it also seems reasonable to
include air-conditioning and pools.  We deliberately left out the number of
bathrooms, the size of the garage, and the size of the lot.  While price seems
to be linearly related to the number of bedrooms, we include it as a factor,
both to check that, and to get three distinct slopes for price on finished area
as quality varies.

```{r, include=FALSE}
initial_mdl <- lm(Price ~ Quality + Sqft*factor(Bedroom)
                  + YearBuild*AdjHighway + Airconditioning + Pool,
                  data=real_estate)
```

This initial model has a root-mean-squared error of \$ $\pm
`r signif(summary(initial_mdl)$sigma, 3)`$, which is not shabby when the median
house price is \$ $`r signif(median(real_estate$Price),3)`$.  Before passing to
issues of model selection, however, such as whether all the interactions are
necessary, whether discrete variables might be usefully recoded, etc., let's
look at the diagnostic plots.

The first thing to say is that the distribution of the residuals doesn't
look very Gaussian, and a Box-Cox transformation suggests the un-intuitive,
indeed  un-interpretable, transformation $1/\sqrt[3]{Y}$.


```{r,qq-and-boxcox-plots-gg, eval=FALSE, out.height="0.2\\textheight", fig.align="center", fig.cap="Q-Q plot of the standardized residuals (left) and Box-Cox plot (right)"}
qq = ggplot(data.frame(residuals = rstandard(initial_mdl)), aes(sample=residuals)) + 
  geom_qq(color="purple") + geom_qq_line(color="orange",size=2) + theme_cowplot()
bc = boxcox(initial.mdl, plotit = FALSE)
ran = bc$x[bc$y > max(bc$y) - 0.5*qchisq(.95,1)]
bcplot = ggplot(data.frame(loglikelihood = bc$y, lambda = bc$x), aes(lambda, loglikelihood)) +
  geom_line(color="purple") + geom_vline(xintercept = bc$x[which.max(bc$y)], color="orange", size=2) +
  geom_vline(xintercept = c(min(ran),max(ran)), color="orange", linetype="dotted") + theme_cowplot()
plot_grid(qq, bcplot)
```


Clients who ask for a model of prices are rarely happy with models for the
inverse cubic roots of prices, so we must be doing something wrong.  Examining
plots of residuals versus predictors suggests that lot size matters after all,
at least for big lots.  The plots also suggest that houses built after $\approx
1980$ are worth more than the model anticipates.  The distributions of
residuals conditional on discrete predictors, however, actually look mostly
homogeneous.

```{r, include=FALSE}
# Function to plot residuals and fitted values vs. predictors
# Note the similarity to the function from hw1
# Inputs: 
#     mdl - a fitted lm model; 
#     dataframe - the name of the data frame; 
#     pred_list - quoted names of continuous predictors
# Output: returns a ggplot object
resid_vs_pred <- function(mdl, dataframe, pred_list) {
    df = dataframe[pred_list]
    df$resids = residuals(mdl) # how do you get residuals?
    df$fitted_values = fitted(mdl) # how do you get the fitted values?
    preds_resids = df %>% 
      pivot_longer(-resids, names_to = 'predictor', values_to = 'value')
    
    ggplot(preds_resids, aes(x=value, y=resids)) + geom_point() +
      geom_smooth() + facet_wrap(~predictor, scales = 'free_x') +
      geom_hline(yintercept = 0, color="red") +
      geom_hline(yintercept = c(2*sd(df$resids),-2*sd(df$resids)),
                 color="grey", linetype="dashed") +
      theme_cowplot(8) + xlab('')
}
```

```{r, diagnostic-plots, fig.height=6, fig.cap=" Residuals versus fitted values and continuous predictors, and versus the discrete predictors.  Grey lines are smoothing splines; dotted lines indicate plus/minus 2 standard deviations, either constant (red) or from a spline smoothing of the squared residuals (grey)."}
cont_resids = resid_vs_pred(
  initial_mdl, real_estate, c("Sqft","YearBuild"))
bxplt_dat = real_estate %>% 
  select(Bedroom, Bathroom, Garage, Highway, Quality, AC) %>%
  mutate_at(vars(Bedroom, Bathroom, Garage), as.factor)
bxplt_dat$residuals = rstandard(initial_mdl)
bxplt_dat = bxplt_dat %>% pivot_longer(-residuals)
bxplts = ggplot(bxplt_dat, aes(y=residuals, x=value, fill=value)) +
  geom_boxplot() + facet_wrap(~name, scales = "free_x") +
  scale_fill_viridis_d() +
  theme_cowplot(8) + theme(legend.position = 'none') + xlab('')
plot_grid(cont_resids, bxplts, nrow = 2, rel_heights = c(1.25,2))
```



## Outliers

In addition to the house with no bedrooms or bathrooms, examination of Cook's
distance shows two houses with exceptional influence over the model.

```{r, cooks-plot, out.height="0.25\\textheight", fig.align="center", fig.cap="Cook's distance for each data point: extremely influential points are flagged in red."}
ggplot(
  tibble(cook = cooks.distance(initial_mdl),
         bad_houses = cook > 0.1,
         price = real_estate$Price),
  aes(price/10000, cook, color=bad_houses)) + geom_point() +
  theme_cowplot() + ylab("Cook's Distance") +
  theme(legend.position = c(.75,.75)) +
  scale_color_manual(values=c("purple","orange")) +
  labs(color = "Bad houses") + xlab("price ($10000s)")
bad_houses <- which(cooks.distance(initial_mdl)>0.1)
```

On examination, these are quite weird: small in area, fairly cheap, but heavy
on bedrooms.  These look more like rental properties than residences.  Checking
the pairs plot again shows no other such anomalies, so we delete them but
leave the rest alone.  Re-doing the other diagnostic plots shows little
over-all change, however (figures omitted).
```{r}
kable(real_estate[bad_houses,c("Price","Sqft","Bedroom","Bathroom")])
```

```{r}
real_estate <- real_estate[-bad_houses,]
```

